Classical Planning Algorithms ¡ on the Atari Video Games

### 2014-11-18 icaps15-paper/paper.tex
Table extracted from main text

\begin{abstract}
\begin{quote}
The Atari  2600 games supported in  the Arcade Learning Environment (Bellemare et. al. 2013)
all feature a known initial (RAM) state and actions that have deterministic effects. Classical planners, however,
cannot be used  for selecting actions  for  two reasons: first,  no  compact PDDL-model of  the games is given,
and  more importantly,  the action effects and goals are  not known \emph{a priori}. Moreover, in these games there is usually
no set of goals to be achieved but rewards to be collected. These features do not preclude the use of   classical  algorithms like
breadth-first search or Dijkstra's algorithm,  but    these methods are not effective over  large state spaces. We thus turn to a
different class of classical planning algorithms  introduced recently  that perform a \emph{structured exploration} of the state space; namely,
like breadth-first search and Dijkstra's  algorithm they are ``blind'' and hence  do not require prior knowledge of state transitions,
costs (rewards) or goals, and yet, like  heuristic search algorithms, they have been shown to be   effective for solving   problems over huge
state spaces. The simplest such algorithm, called Iterated Width or \IW, consists of  a sequence of calls  $\IW(1)$, $\IW(2)$, \ldots,
$\IW(k)$ where  $\IW(i)$ is  a breadth-first search in which a state is  pruned when it is not the
first state in the search  to make true some set of $i$ atoms. 
The  empirical results suggest that the performance of \IW\ with the  $k$ parameter fixed to $1$, 
i.e., $\IW(1)$, approaches the performance of the state of the art represented by 
 \UCT, and that a simple best-first variation  of \IW\  that combines exploration 
and exploitation is even more competitive.

\end{quote}
\end{abstract}

@@

The number of nodes \emph{generated} by $\IW(1)$ is $n \times D \times b$ in the worst case,
where $n$ is the number of variables, $D$ is the size of their domains, and $b$ is the number of
actions per  state.  This same number in breadth-first search is not linear in $n$
but exponential. For the Atari games,  $n=128$, $D=256$, and $b=18$, so that the product is
equal to $589,824$, which is large but feasible. On the other hand, the number
of nodes generated by $\IW(2)$ in the worst case is  $(n \times D)^2 \times b$, which is
equal to $1,073,741,824$ which is too large, forcing us to consider only a fraction of
such states. For classical planning problems, the growth in the number of nodes from
$\IW(1)$ to $\IW(2)$ is not that large, as the variables are boolean. Indeed, we could have
taken the state vector for the Atari games as a vector of 1024 boolean variables,
and apply these algorithms to that representation. The results would be different.
In such a case, $\IW(1)$ and $\IW(2)$  would generate $n' \times D' \times b$ and
$(n' \times D')^2 \times b$ states, which for $n'=1024$ and $D=2$
represent  $36,864$ and $7,5497,472$ states respectively.
These are feasible numbers, but we haven't used  this representation 
under the assumption that the correlations among the bits in each one of
the 128 words of the state vectors are meaningful.
In summary, from the basic \IW\ algorithm we are testing only the first
call $\IW(1)$.

@@

For  the experiments below, we added two simple variations to  $\IW(1)$ and \BFS.
First, in the breadth-first search underlying $\IW(1)$, we generate the children in
random order. This makes the executions that results from the
$\IW(1)$ lookahead less susceptible to fall into loops; a potential
problem in local search algorithms with no memory or learning.
Second, a discount factor $\gamma=0.995$ is used in both algorithms
for discounting future rewards like in \UCT.
For this, each state $s$ keeps its depth $d(s)$ in the search tree,
and if state s' is the child of state $s$ and action $a$,
$R(s')$ is set to $R(s) + \gamma^d(s´) r(a,s)$.
The  discount factor  results in a slight
preference for rewards that can be reached earlier,
which is a  reasonable heuristic in  on-line settings
based on lookahead searches.

@@

\begin{table*}
\centering
% \addtolength{\tabcolsep}{-2pt}
% {\scriptsize
{\small
\begin{tabular}{@{}l|rr|rr|r|r@{}}
%% table header
% \hline\\[-3ex]
%\hline
\multicolumn{1}{c|}{}& \multicolumn{2}{c|}{$\IW(1)$}& \multicolumn{2}{c|}{\BFS} & \multicolumn{1}{c|}{\BRFS} &  \multicolumn{1}{c}{\UCT}
% \\[.2ex]
\\
% \hline\\\\[-3ex]
\hline
%% table body
\multicolumn{1}{c|}{Game}  &
% \multicolumn{1}{c|}{Avg. Score} &
% \multicolumn{1}{c}{Avg. Score} & \multicolumn{1}{c|}{Avg. Time}  &
% \multicolumn{1}{c}{Avg. Score} & \multicolumn{1}{c|}{Avg. Time}  &
% \multicolumn{1}{c}{Avg. Score} & \multicolumn{1}{c|}{Avg. Time}  &
% \multicolumn{1}{c}{Avg. Score} & \multicolumn{1}{c}{Avg. Time}\\[.2ex]
 \multicolumn{1}{c}{Score} & \multicolumn{1}{c|}{Time}  &
 \multicolumn{1}{c}{Score} & \multicolumn{1}{c|}{Time}  &
 \multicolumn{1}{c|}{Score} &
 \multicolumn{1}{c}{Score} 
% \\[.2ex]
\\[.2ex]
% \hline\\
\hline

\propername{Assault} & 806.4 & 18.21 & \textbf{852.6} & 25.11 & 413.7 & 1512.2 \\ \hline
\propername{Asterix} & 153400 & 24.33 & 77200 & 26.58 & 2135.7 & \textbf{290700} \\ \hline
\propername{Asteroids} & \textbf{51338} & 65.92 & 22168 & 64.53 & 3127.4 & 4660.6 \\ \hline
\propername{Atlantis} & 159420 & 12.8 & \multicolumn{1}{l}{--} & \multicolumn{1}{l|}{--} & 30460 & \textbf{193858} \\ \hline
\propername{Bank\_heist} & \textbf{717} & 39.29 & 362 & 63.83 & 21.5 & 497.8 \\ \hline
\propername{Battle\_zone} & 11600 & 85.96 & \multicolumn{1}{l}{--} & \multicolumn{1}{l|}{--} & 6312.5 & \textbf{70333.3} \\ \hline
\propername{Beam\_rider} & 9108 & 22.53 & \textbf{9297.6} & 29.03 & 693.5 & 6624.6 \\ \hline
\propername{Berzerk} & \textbf{2096} & 58 & 802 & 73.47 & 195 & 553.5 \\ \hline
\propername{Bowling} & \textbf{68.6} & 9.84 & 50.4 & 59.52 & 25.5 & 25.1 \\ \hline
\propername{Boxing} & 100 & 15.2 & 100 & 22.3 & 100 & 100 \\ \hline
\propername{Breakout} & 384.2 & 4.22 & \textbf{771.6} & 39.47 & 1.1 & 364.4 \\ \hline
\propername{Centipede} & 99207.2 & 39.14 & 93067 & 65.49 & \textbf{125123} & 110422 \\ \hline
\propername{Chopper\_command} & 2100 & 41.06 & 3800 & 40.94 & 1827.3 & \textbf{34018.8} \\ \hline
\propername{Crazy\_climber} & 36160 & 3.61 & 36940 & 58.12 & 37110 & \textbf{98172.2} \\ \hline
\propername{Enduro} & \textbf{500} & 65.69 & 359.4 & 37.66 & 0.6 & 286.3 \\ \hline
\propername{Freeway} & \textbf{30.6} & 31.69 & 22.6 & 60.77 & 0 & 0.4 \\ \hline
\propername{Frostbite} & \textbf{3582} & 12.25 & \multicolumn{1}{l}{--} & \multicolumn{1}{l|}{--} & 137.2 & 270.5 \\ \hline
\propername{Gravitar} & 3920 & 61.85 & \textbf{5980} & 62.45 & 395 & 2850 \\ \hline
\propername{Hero} & \textbf{12887} & 36.63 & 7096.6 & 51.83 & 1323.8 & 12859.5 \\ \hline
\propername{Jamesbond} & \multicolumn{1}{l}{--} & \multicolumn{1}{l|}{--} & \textbf{8550} & 25.68 & 25 & 330 \\ \hline
\propername{Kangaroo} & \textbf{5700} & 7.9 & 5320 & 31 & 90 & 1990 \\ \hline
\propername{Krull} & \multicolumn{1}{l}{--} & \multicolumn{1}{l|}{--} & \textbf{5124} & 36.82 & 3089.2 & 5037 \\ \hline
\propername{Montezuma\_revenge} & 0 & 17.36 & \textbf{80} & 34.72 & 0 & 0 \\ \hline
\propername{Ms\_pacman} & \multicolumn{1}{l}{--} & \multicolumn{1}{l|}{--} & 14781 & 23.97 & 1708.5 & \textbf{22336} \\ \hline
\propername{Pong} & \textbf{21} & 16.67 & \textbf{21} & 34.92 & -20.7 & \textbf{21} \\ \hline
\propername{Pooyan} & 11225 & 7.97 & \multicolumn{1}{l}{--} & \multicolumn{1}{l|}{--} & 909.7 & \textbf{17763.4} \\ \hline
\propername{Qbert} & 3705 & 11.07 & 11680 & 34.61 & 132.8 & \textbf{17343.4} \\ \hline
\propername{Riverraid} & \multicolumn{1}{l}{--} & \multicolumn{1}{l|}{--} & \textbf{5062} & 36.53 & 2178.5 & 4449 \\ \hline
\propername{Seaquest} & \textbf{14272} & 24.91 & 6138 & 33.38 & 288 & 5132.4 \\ \hline
\propername{Space\_invaders} & 2877 & 21.37 & \textbf{3974} & 34.21 & 112.2 & 2718 \\ \hline
\propername{Star\_gunner} & 1540 & 18.89 & \textbf{4660} & 17.62 & 1345 & 1207.1 \\ \hline
\propername{Tennis} & \textbf{23.8} & 21.05 & 23.6 & 36.33 & -23.8 & 2.8 \\ \hline
\propername{Venture} & 775 & 23.41 & \textbf{1080} & 27.78 & 0 & 0 \\ \hline
\propername{Video\_pinball} & \multicolumn{1}{l}{--} & \multicolumn{1}{l|}{--} & 62074.6 & 43.13 & 55567.3 & \textbf{254748} \\ \hline
\propername{Zaxxon} & \multicolumn{1}{l}{--} & \multicolumn{1}{l|}{--} & 15680 & 31.01 & 0 & \textbf{22610} \\ \hline
\multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} \\ \hline
Times Best (35 games)& \multicolumn{2}{c|}{12} & \multicolumn{2}{c|}{12} & 1 & 11 \\ \hline
Times Better than UCT (35 games) & \multicolumn{2}{c|}{23} & \multicolumn{2}{c|}{22} & 1  & -- \\ \hline

\end{tabular}
}

 \caption{\small Performance that results from random policy and various  lookahead algorithms in 35 Atari games.
The algorithms  are \BRFS, $\IW(1)$, \BFS, and \UCT. Averages  over 5 runs shown with ``random'' seeds
$0$ to $4$, and maximum episode duration of $18,000$  frames. Every algorithm limited to a budget of
150,000 simulated frames, except for UCT limited to 500 rollouts, 300 frames deep instead. Average scores and CPU times per action selection shown.
Performance of UCT taken from \cite{bellemare:jair2013}.
Numbers  in bold show best performer in terms of average score.
--Domains containing (--) are still computing,  each run can take 24 hours of computation. All domains from \cite{bellemare:jair2013}  will be ready for the camera ready--. }
\label{performance}

\end{table*}
 We tested $\IW(1)$ and \BFS\ over $35$ games: those considered in
\cite{bellemare:jair2013}, abbreviated as BNVB \footnote{The full set
  of games is made of 55 games, 5 runs each. As each run takes around
  24 hours of CPU time, we luck the resources to submit the full table
  by the submission date, but the full table will be ready for the
  camera ready version.}. In Table~\ref{performance}, the
performance of $\IW(1)$ and \BFS\
% , characterized by the average score obtained over all episodes, the standard deviation
% of the score obtained and time (in seconds) per lookahead,
is compared with that of \UCT\ and breadth-first search (\BRFS).
The discount factor used by all the algorithms is $\gamma=0.995$ and the average scores
reported for \BRFS\ and \UCT\ in  BNVB  are included for reference. 
Our experimental setup  follows the one in BNVB with one exception: a maximum budget of  $150,000$ simulated frames 
is applied to \textit{all the algorithms}, including  \UCT\ that is limited to  $500$ base policy rollouts of depth $300$.
% \footnote{In \UCT,  $500$ base policy rollouts
% of depth $300$ is equivalent to a budget of $150,000$ simulated frames.}.
The bound on the number of simulated frames is like a bound on
lookahead time, as most of the time in the lookahead search is spent
in the calls to the emulator for computing the next RAM state.  Also,
as in BNVB, all of the algorithms reuse the subtree from the previous
lookahead that is rooted in the selected child, deleting its siblings
and their descendants. More precisely, no calls to the emulator are
done for transitions that are cached in that subtree. As a result the
number of nodes in the lookahead tree may be larger actually than the
budget. In addition, in \IW, the states that are ``reused'' from the
previous searches are ignored in the computation of the novelty of new
states so that more states can escape pruning. Otherwise, \IW\ often
doesn't use the full budget. This is not needed in \BFS\ which does no
pruning.  $\IW(1)$ and \BFS\ have been limited the search to a depth of
$m=1500$ frames and up to $m=150,000$ frames per root branch. This is
to avoid the search for going too deep or being too committed to a
single root action. Similarly UCT defines a limit of $m=300$ frames
for each rollout. Note that UCT is more likely to have a performance
loss if the depth is increased, as the number of rollouts would
decrease in order to maintain the same budget of $150,000$ simulated
frames. The last difference inherited from BNVB between all algorithms
and UCT, is that UCT selects an action randomly in the rollout every
single frame, while the others select an action every 5 frames, and
the remain ones are $no-op$, as is done in the simulation
environment. Therefore, a node in all algorithms represent $5$ frames,
but in UCT represents $1$ frame. As a result, a UCT rollout of depth
$300$ frames involve the selection of 300 actions, decreasing the
budget of simulated frames by $300$, while the other algorithms need a
branch of $1500$ frames to select $300$ actions, decreasing the budget
by $1500$ frames instead.

Table~\ref{performance} shows that $\IW(1)$ and \BFS\ both outperform
\BRFS, which rarely collects any reward in many domains, as the
depth of the \BRFS\ search tree results in a lookahead of $0.3$
seconds ($20$ frames or $4$ nodes deep). On the other hand, both
$\IW(1)$ and \BFS\ reach states that are up to $350$--$1500$ frames
deep ($70$--$260$ nodes or $6$--$22$ seconds), even if $\IW(1)$
doesn't always use all the simulation frames allocated due to its
agressive pruning. This behaviour can be observed in games such as
\Breakout, \CrazyClimber, \Kangaroo, and \Pooyan, where the average
CPU time for each lookahead is up to $10$ times faster than \BFS. Note
that computation time for \UCT\ and \BRFS\ is similar to \BFS, as the
most expensive computation is to generate a frame through the
simulator and all algorithms use always the full budget.

Interestingly, in terms of average scores out of $35$ games, simple $\IW(1)$ outperforms
\BFS\ and \UCT\ in $18$ and $23$ games respectively,
most often with lower average times per call. \BFS\ outscores \UCT\ in
$22$ of the games \footnote{Note that both \IW\ and \BFS\ may improve, as 6 and 4 games  respectively are not reported in the table
  yet.}. As mentioned in the
previous section, the order in which $\IW(1)$ and \BFS\ generate the
children of an expanded node is randomized, something which has a
potential effect on the novelty measures of nodes, and hence in the
search graph. We found that this randomization helps, in particular in
\Qbert, where not randomizing results in $\IW(1)$ being trapped into
loops sometimes.

Likewise, $\IW(1)$ outperforms the best learning algorithm in
\cite{deep-mind-atari} in $7$ out of the $7$ games considered, and
\BFS\ does so in $6$ of the $7$ games.

We also note that in \Freeway, \Berzerk, \Bowling, \Montezuma,
\StarGunner, and \Venture, both $\IW(1)$ and \BFS\ attain a better
score than the baseline semi-random algorithm \emph{Perturb} in
\cite{bellemare:jair2013}, which outperforms both \UCT\ and the
reinforcement learning methods. \emph{Perturb} performance helps
understanding these domains. \emph{Perturb} selects a fixed action with probability
0.95 and otherwise acts uniformly randomly, and the best run out of 10
episodes is reported in BNVB. Note that all the other algorithms
report average score instead. The behaviour of \emph{Perturb} hinder
the nature of these domains: rewards are very deep, but tipically
repeating the same action may lead to some reward.

Consider \Freeway, where a chicken has to run across a ten lane
highway filled with traffic. Every time it gets across, we
receive a point, i.e. a reward of 1. If the chicken is hit by a car,
the chicken is forced back. 12 out of 18 possible actions have an
effect: move up ( up-right, up-left, up-fire, up-right-fire,
up-left-fire, all have same effect as up ) or move down ( down-right,
down-left, down-fire, down-right-fire, down-left-fire, all have same
effect as down ), the remaining 6 actions are like no-ops. It's easy
to understand that \emph{Perturb} would achieve a high score if the
selected action is move up with 0.95 probability. In the following link
\url{http://www.youtube.com/playlist?list=PLXpQcXUQ_CwenUazUivhXyYvjuS6KQOI0}
we can observe videos of \IW\ and \UCT\ playing \Freeway. As reported
in Table~\ref{performance}, \UCT\ does not manage to take any chicken
across the highway, while \IW\ is the best performer out of all tested
algorithms. The main reason why \UCT\ performs so erratically is that
the rollouts do not manage to collect any reward, as there is a
$3^{-1}$ probability to choose any move up action, and given that the
depth of the closest reward is about 200 frames deep, the probability
of a rollout achieveing any reward is $3^{-200}$, which implies
choosing 200 move-up actions. On the other hand, \IW\ manages to find
rewards without exhausting the full budget. The main reason that
explains this behaviour is that at most only 3 generated states from
the same parent node will pass the novelty test, as in any expanded
node there is at most 3 different new possible bytes in the RAM state
resulting from moving up, down, or do-nothing.

Another challenging game is \Montezuma. In this game rewards are very
sparse, deep, and most of the actions lead to loosing a life, which
only becomes effective after hundreds of frames, i.e. up-left may
enforce jumping and falling until we loose a life. All algorithms achieve 0 score, wheter RL-based or Planning-based. But, as
shown in Table~\ref{performance}, \BFS\ achieves an average score of
80, which is still a very low score, as the first possible reward
increases the score in 100. The best run by \BFS\ achieves a score of
400, pointing out that \BFS\ still is not able to consistently find rewards. One way to improve the performance of any algorithm in \Montezuma\ is to encode the lost of a life as a negative reward. Capturing the lost of lifes in the reward function is relevant to many games where not loosing a life is important in later stages of the game. Currently,
rewards are encoded directly as the change of score, which does not distinguish between the benefits of keeping lifes. This
beahaviour is also observed in other games like \Breakout\ and \SpaceInvaders, where \BFS\
is the best performer by far, although initially it looses all lifes but 1 and then plays extremely well. Videos on the link above illustrate
this behaviour.

We do not report the performance of $\IW(2)$ because  in our
preliminary tests and according to the discussion in the previous section,
it doesn't  appear to improve much on  \BRFS, even if  it results in a lookahead
that is $5$ times deeper,  still far too shallow  to compete with the other
planning algorithms.

@@

The notions of width and the \IW\  algorithm guarantee that  states with low width will be generated
in low polynomial time through \emph{shortest} paths. In the presence of rewards like the Atari games,
however, the  interest is not  in the shortest paths but in the \emph{best} paths; i.e, the paths with
maximum reward.  \IW\ may actually  fail to find such paths even when calling \IW($k$) with a high 
$k$ parameter. Optimality  could be  achieved then  by replacing the breadth-first search
underlying \IW($k$) by Dijkstra's algorithm yet such a move would make the relation between
\IW\ and the notion of width less clear. Another option is to change  \IW\ to comply with 
a different property; namely, that if there is a ``rewarding'' path made up of states of 
low width, then \IW\ will find such paths or better ones in time that is exponential 
in their width. For this actually a simple change in \IW\ suffices: when generating a state 
$s$ that is a \emph{duplicate} of a state $s'$ that has been previously generated and not pruned, 
$s'$ is replaced by $s$ if $R(s) > R(s')$. Moreover in such a case, the change in reward 
must be propagated to the descendants of $R(s')$ that are in memory. This is similar to the change required 
in A* for preserving optimality when moving from consistent to inconsistent heuristics \cite{pearl:heuristics}. 
The alternative is to ``reopen'' such nodes. 
%The same change is actually in \BFS\ to ensure that, 
%if given enough time, \BFS\ will actually find the optimal paths.
Changing \BFS\ to account for such paths guarantees that, if given enough time, \BFS\ will actually find
the optimal paths. 
The code used for \IW\ and \BFS\ in the
experiments above does not implement this change, however, as the overhead involved in checking for duplicates 
in some test cases did not appear to pay off. More experimentation is probably needed.

>>>>

\begin{abstract}
\begin{quote}
The Atari  2600 games supported in  the Arcade Learning Environment (Bellemare et. al. 2013)
all feature a known initial (RAM) state and actions that have deterministic effects. Classical planners, however,
cannot be used  for selecting actions  for  two reasons: first,  no  compact PDDL-model of  the games is given,
and  more importantly,  the action effects and goals are  not known \emph{a priori}. Moreover, in these games there is usually
no set of goals to be achieved but rewards to be collected. These features do not preclude the use of   classical  algorithms like
breadth-first search or Dijkstra's algorithm,  but    these methods are not effective over  large state spaces. We thus turn to a
different class of classical planning algorithms  introduced recently  that perform a \emph{structured exploration} of the state space; namely,
like breadth-first search and Dijkstra's  algorithm they are ``blind'' and hence  do not require prior knowledge of state transitions,
costs (rewards) or goals, and yet, like  heuristic search algorithms, they have been shown to be   effective for solving   problems over huge
state spaces. The simplest such algorithm, called Iterated Width or \IW, consists of  a sequence of calls  $\IW(1)$, $\IW(2)$, \ldots,
$\IW(k)$ where  $\IW(i)$ is  a breadth-first search in which a state is  pruned when it is not the
first state in the search  to make true some subset of $i$ atoms. 
The  empirical results over 35 games suggest that the performance 
of \IW\ with the  $k$ parameter fixed to $1$, i.e., $\IW(1)$, is at the level of the state of the art represented by 
\UCT.  A simple best-first variation  of \IW\  that combines exploration and exploitation proves
to be very competitive as well. 
\end{quote}
\end{abstract}

@@

The number of nodes \emph{generated} by $\IW(1)$ is $n \times D \times b$ in the worst case,
where $n$ is the number of variables, $D$ is the size of their domains, and $b$ is the number of
actions per  state.  This same number in breadth-first search is not linear in $n$
but exponential. For the Atari games,  $n=128$, $D=256$, and $b=18$, so that the product is
equal to $589,824$, which is large but feasible. On the other hand, the number
of nodes generated by $\IW(2)$ in the worst case is  $(n \times D)^2 \times b$, which is
equal to $1,073,741,824$ which is too large, forcing us to consider only a tiny fraction of
such states. For classical planning problems, the growth in the number of nodes from
$\IW(1)$ to $\IW(2)$ is not that large, as the variables are boolean. Indeed, we could have
taken the state vector for the Atari games as a vector of 1024 boolean variables,
and apply these algorithms to that representation. The results would be different.
In such a case, $\IW(1)$ and $\IW(2)$  would generate $n' \times D' \times b$ and
$(n' \times D')^2 \times b$ states, which for $n'=1024$ and $D=2$
represent  $36,864$ and $7,5497,472$ states respectively.
These are feasible numbers, but we haven't used  this representation 
under the assumption that the correlations among the bits in each one of
the 128 words of the state vectors are meaningful.
In summary, from the basic \IW\ algorithm we are testing only the first
call $\IW(1)$.

@@

For  the experiments below, we added two simple variations to  $\IW(1)$ and \BFS.
First, in the breadth-first search underlying $\IW(1)$, we generate the children in
random order. This makes the executions that results from the
$\IW(1)$ lookahead less susceptible to fall into loops; a potential
problem in local search algorithms with no memory or learning.
Second, a discount factor $\gamma=0.995$ is used in both algorithms
for discounting future rewards like in \UCT.
For this, each state $s$ keeps its depth $d(s)$ in the search tree,
and if state s' is the child of state $s$ and action $a$,
$R(s')$ is set to $R(s) + \gamma^{d(s)+1} r(a,s)$.
The  discount factor  results in a slight
preference for rewards that can be reached earlier,
which is a  reasonable heuristic in  on-line settings
based on lookahead searches.

@@

\input{games-table}
We tested \IW(1) and \BFS\ over $35$ of the 55 games considered in
\cite{bellemare:jair2013}, from now on  abbreviated as BNVB.\footnote{
Entries marked as  ``--'' in the table are   still running at the time of submission. 
Each run can take up to 24 hours of CPU time and there are 5 runs per game and algorithm. 
Eventually all of  the  55 games will  be included in the evaluation.}
Table~\ref{performance} shows the performance of $\IW(1)$ and \BFS\
% , characterized by the average score obtained over all episodes, the standard deviation
% of the score obtained and time (in seconds) per lookahead,
in comparison  with breadth-first search (\BRFS) and \UCT.
Videos of selected games played by \IW(1), \BFS, and \UCT\ can be 
seen in Youtube.\footnote{\url{http://www.youtube.com/playlist?list=PLXpQcXUQ_CwenUazUivhXyYvjuS6KQOI0}.}
The discount factor used by  all the algorithms is $\gamma=0.995$.
The scores for \BRFS\ and \UCT\ are from BNVB.  
Our experimental setup  follows theirs except that a maximum budget of  $150,000$ simulated frames 
is applied to \textit{all the algorithms}. \UCT\ uses this budget by running
 $500$   rollouts of depth $300$. % \footnote{In \UCT,  $500$ base policy rollouts
% of depth $300$ is equivalent to a budget of $150,000$ simulated frames.}.
The bound on the number of simulated frames is like a bound on
lookahead time, as most of the time in the lookahead  is spent
in calls to the emulator for computing the next RAM state. 
This is why the average time per action is similar to all the algorithms
except \IW(1), that due to its agressive pruning does not always use
the full budget and takes less time per action on average.

Also,  as reported by  BNVB, all of the algorithms 
reuse the subtree from the previous lookahead that is rooted in the selected 
child, deleting its siblings and their descendants. More precisely, no calls 
to the emulator are done for transitions that are cached in that subtree, 
and such reused frames are not discounted from the budget that is thus a bound 
on the number of ``new'' frames per lookahead. In addition, in \IW(1), the states that 
are ``reused'' from the previous searches are ignored in the computation of the novelty of 
new states so that more states can escape pruning. Otherwise, \IW(1) often
uses a fraction of the  budget.  This is not needed in \BFS\ which does 
no pruning.  \IW(1) and \BFS\ are  limited to  search up to a depth of
$m=1500$ frames and up to $m=150,000$ frames per root branch. This is
to avoid the search from  going too deep or being too committed to a
single root action. 
% Similarly UCT defines a limit of $m=300$ frames for each rollout. 
% Note that UCT is more likely to have a performance
% loss if the depth is increased, as the number of rollouts would
% decrease in order to maintain the same budget of $150,000$ simulated
% frames.

Last, \IW(1) and \BFS\ select an action every 5 frames, while UCT  selects 
an action in every frame.  This means that in order to explore a branch 
300 frames deep, UCT gets to choose 300 actions, while \IW(1) and \BFS\ 
get to choose 60 actions, both however using the same 300 frames from the
budget. For this, we  followed the setup  of \BRFS\ in BNVB 
that also selects  actions every 5 frames. Since the lookahead budget 
is given by a maximum number of (new) frames, and the time is mostly taken by 
calls to the emulator, this may not be the best choice  for \IW(1)\ 
and \BFS\ that are therefore not exploiting all the options afforded
by the budget. We hope to fix this gap in the future.

\Omit{
The last difference inherited from BNVB between all algorithms
and UCT, is that UCT selects an action randomly in the rollout every
single frame, while the others select an action every 5 frames, and
the remain ones are $no-op$, as is done in the simulation
environment. Therefore, a node in all algorithms represent $5$ frames,
but in UCT represents $1$ frame. As a result, a UCT rollout of depth
$300$ frames involve the selection of 300 actions, decreasing the
budget of simulated frames by $300$, while the other algorithms need a
branch of $1500$ frames to select $300$ actions, decreasing the budget
by $1500$ frames instead.
}

Table~\ref{performance} shows that both $\IW(1)$ and \BFS\  outperform
\BRFS, which rarely collects any reward in many domains, as the
depth of the \BRFS\ search tree results in a lookahead of $0.3$
seconds ($20$ frames or $4$ nodes deep). On the other hand, both
$\IW(1)$ and \BFS\ reach states that are up to $350$--$1500$ frames
deep ($70$--$260$ nodes or $6$--$22$ seconds), even if $\IW(1)$
doesn't always use all the simulation frames allocated due to its
pruning. This behaviour can be observed in games such as
\Breakout, \CrazyClimber, \Kangaroo, and \Pooyan, where the average
CPU time for each lookahead is up to $10$ times faster than \BFS. 
Computation time for \UCT\ and \BRFS\ are  similar to \BFS, as the
most expensive part of the computation is the  generation of 
frames through the simulator,  and these three  algorithms 
always use  the full budget.

More interestingly,  \IW(1) outscores \UCT\  in 17 of the 35 games,
while \BFS\ outscores \UCT\ in 18.\footnote{These numbers can actually
improve further as 6  games for \IW(1) and 4 games for \BFS\ are still running.}
On the other, \UCT\ does better than \IW(1) and \BFS\ in 9 and 11 games respectively.
The relative performance between \IW(1)\ and \BFS\ is closer with \IW(1) being  the
best of the two in 12 games, and  \BFS\  in 11. In terms of the number of games 
where an algorithm is the overall winner, the situation is  even: 
\IW(1) is the overall best in 12 games, \BFS\ in 11 games, and \UCT\ in 12 games. 
Also, \BRFS\ is best in 1 game (Centipede), while the other three algorithms
are tied in another game (\Pong).

\Omit{As mentioned in the
previous section, the order in which $\IW(1)$ and \BFS\ generate the
children of an expanded node is randomized, something which has a
potential effect on the novelty measures of nodes, and hence in the
search graph. We found that this randomization helps, in particular in
\Qbert, where not randomizing results in $\IW(1)$ being trapped into
loops sometimes.}

$\IW(1)$ also outperforms the best learning algorithm in
\cite{deep-mind-atari} in the $7$ games considered, and
\BFS\ does so in $6$ of the $7$ games. Likewise, in \Freeway, \Berzerk, \Bowling, \Montezuma,
\StarGunner, and \Venture, both $\IW(1)$ and \BFS\ attain a better
score than the baseline semi-random algorithm \emph{Perturb} in
\cite{bellemare:jair2013}.
%% to say the 2 lines below must add context: .. eg
%%% ``which in *these games* outperforms both \UCT\ and the reinforcement learning methods.''
%% add it if that's the case ..
\emph{Perturb} is a simple algorithm
that selects  a fixed  action with probability 0.95, and a random
action with probability 0.05.  For \emph{Perturb}, BNVB do not report 
the average score  but the  best score.  \emph{Perturb} manages to do well in domains where rewards
are deep but can be  reached by  repeating the same action.
This is the case of \Freeway, where a chicken has to run across a ten lane
highway filled with traffic. Every time the chicken gets across, there is one
unit of reward. If the chicken is hit by a car, it has to start again. 
In \Freeway, 12 out of 18 possible actions have an effect:
6 actions behave as move-up (up-right, up-left, up-fire, up-right-fire, up-left-fire) , 
6 as  move-down (down-right, down-left, down-fire, down-right-fire, down-left-fire), 
and 6 actions do nothing.  \emph{Perturb}  does well in this domain 
when the selected fixed action is move-up.  As seen 
in Table~\ref{performance}, \UCT\ does not manage to take any chicken
across the highway. The reason that  \UCT\ performs so erratically 
and the rollouts do not manage to collect any reward, is that there  is a $1/3$ probability of  
choosing a move up action,  and   200 move up actions are required  in order to collect a reward. 
%% the previous number was not correct; no-op actions do not count, and thus needs
%%% 200 more move up actions than move down actions, although this is not accurate either
%%%% as move down actions that take chicken below the screen count as no-ops too ..
\IW\ does not have this limitation and is best in \Freeway.

%%
%%% Sobre texto below; no me convence; no sabemos como cambia el RAM aunque podamos suponer
%%% obvio: si un byte codifica donde esta el chicken en la pantalla .. tendriamos una explicación
%% On the other hand, \IW\ manages to find rewards without exhausting the  budget. The main reason that
%% explains this behaviour is that at most only 3 generated states from
%% the same parent node will pass the novelty test, as in any expanded
%% node there is at most 3 different new possible bytes in the RAM state
%% resulting from moving up, down, or do-nothing.

Another challenging game is \Montezuma. In this game rewards are very
sparse, deep, and most of the actions lead to losing a  life with
no immediate penalty or consequence.  All algorithms achieve 0 score, 
except for \BFS\ that achieves an average score of 80, and a score
of 400 in one of the runs. This means that even \BFS\  is not able to 
consistently find rewards in this game. This game and several other games 
like \Breakout, and \SpaceInvaders\ could be much simpler by adding
negative rewards for losing a life. We have indeed observed that
our planning algorithms don't care much about losing lifes
until there is just one life left, when they play much better.
This can be seen in the videos mentioned above, and points
to a simple form of learning that would be useful both
for planners and reinforcement learning approaches.

We are not  reporting  the performance of $\IW(k)$ with parameter $k=2$ because  in our
preliminary tests and according to the discussion in the previous section,
it doesn't  appear to improve much on  \BRFS, even if  it results in a lookahead
that is $5$ times deeper,  but  still too shallow  to compete with the other
planning algorithms.

@@

The notions of width and the \IW\  algorithm guarantee that  states with low width will be generated
in low polynomial time through \emph{shortest} paths. In the presence of rewards like the Atari games,
however, the  interest is not  in the shortest paths but in the \emph{best} paths; i.e, the paths with
maximum reward.  \IW\ may actually  fail to find such paths even when calling \IW($k$) with a high 
$k$ parameter. Optimality  could be  achieved then  by replacing the breadth-first search
underlying \IW($k$) by Dijkstra's algorithm yet such a move would make the relation between
\IW\ and the notion of width less clear. A better  option is to change  \IW\ to comply with 
a different property; namely, that if there is a ``rewarding'' path made up of states of 
low width, then \IW\ will find such paths or better ones in time that is exponential 
in their width. For this,  a simple change in \IW\ suffices: when generating a state 
$s$ that is a \emph{duplicate} of a state $s'$ that has been previously generated and not pruned, 
$s'$ is replaced by $s$ if $R(s) > R(s')$. Moreover in such a case, the change in reward 
must be propagated to the descendants of $R(s')$ that are in memory. This is similar to the change required 
in the  A* search algorithm for preserving optimality when moving from consistent to inconsistent heuristics \cite{pearl:heuristics}. 
The alternative is to ``reopen'' such nodes. The same change is actually needed in \BFS\ to ensure that, 
if given enough time, \BFS\ will actually find optimal paths.
% Changing \BFS\ to account for such paths guarantees that, if given enough time, \BFS\ will actually find
% the optimal paths. 
The code used for \IW\ and \BFS\ in the
experiments above does not implement this change  as the overhead involved in checking for duplicates 
in some test cases did not appear to pay off. More experiments  however are  needed to
find out if this is actually the most effective option.
